Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-09 analysis/paper/paper.Rmd
work on discussion

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments, during the presence of the European that hints the emergence of social inequality. Using an EFA-based outline approach, we examine the degree of the ceramic standardization as an indicator of the presence of ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method. There are differences in pottery shapes detected in the indigenous society across different foreign contact phases, including the European in the 17th century and the Chinese in the 19th century. The significant difference especially presented between Chinese contact and either pre-European or post-European (P-value < 0.01). We also observe there is no obvious changes in the clay paste across different phases in relation to the increasing standardization of the pottery shape. More homogeneous and standardized in shape without changes in raw materials reflects the clay source might be the same but a preferred shape was emphasized during the Chinese contact period. This might indicate the appearance of specialized group. The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases. It would need further analysis to understand if the changes in shape related to the function of pot.

>>>>

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments, during the presence of the European that hints the emergence of social inequality. Using an EFA-based outline approach, we examine the degree of the ceramic standardization as an indicator of the presence of ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method. There are differences in pottery shapes detected in the indigenous society across different foreign contact phases, including the European in the 17th century and the Chinese in the 19th century. The significant difference especially presented between Chinese contact and either pre-European or post-European (P-value < 0.01). We also observe there is no obvious changes in the clay paste across different phases in relation to the increasing standardization of the pottery shape. More homogeneous and standardized in shape without changes in raw materials reflects the clay source might be the same but a preferred shape was emphasized during the Chinese contact period. The more standardized shape might indicate the appearance of specialized group. The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases. To figure out whether this change is related to the function of pot, we also attempt to conduct a preliminary lipid analysis for potsherds. However, we did not get useful results due to the low lipid yield for meaningful interpretation.

The result of shape analysis suggests that there might be the emergence of specialized groups after the contact with the Europeans. This supports our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society. This inequality might develop based on the monopolies for trade between a small number of indigenous people and the social organization switch from a more corporate to a more network organization [@Feinman2000].
